Tsbrowse: an interactive browser for Ancestral Recombination Graphs

### 2025-04-28 manuscript/paper.tex
added ORCID details; fixed suppl. fig 1 floating; final readthrough;

\author[1$\ast$]{Savita Karthikeyan}
\author[1]{Ben Jeffery}
\author[1]{Duncan Mbuli-Robertson}
\author[1]{Jerome Kelleher \ORCID{0000-0000-0000-0000}}

@@

\section{Introduction}
Ancestral recombination graphs (ARGs) describe how a set of sample
sequences relate to each other at each position along the genome in a
recombining
species, and are currently the subject of intense research
interest~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
    wong2024general}. ARGs are a fundamental object in population genetics,
and although they have been of theoretical interest for
decades~\citep{Hudson1983,Griffiths1997} it is only with recent breakthroughs
in inference methods~\citep{rasmussen2014genome,
    speidel2019method,kelleher2019inferring,wohns2022unified,zhang2023biobank,
    gunnarsson2024scalable,deng2024robust} that practical application has
become possible. Varied applications have been recently proposed,
such as inferring selection~\citep{stern2019approximate,hejase2022deep} and
the spatial location of genetic ancestors~\citep{osmond2024estimating,deraje2024inferring,grundler2024geographic},
more powerful approaches to quantifying genetic
relatedness~\citep{fan2022genealogical,zhang2023biobank,gunnarsson2024scalable,lehmann2025on}
and other methodological improvements for genome wide association
studies~\citep{nowbandegani2023extremely,link2023tree},
and the development of machine learning methods using inferred
ARGs as input~\citep{hejase2022deep,pearson2023local,korfmann2024simultaneous,
whitehouse2024tree}.
While these developements are exciting, the performance 
of these new methods depends critically on the accuracy of the inferred ARGs.
Although studies benchmarking the various inference methods on 
simulated data have emerged 
\citep{brandt2022evaluation,deng2024robust,peng2024evaluating},
the practicalities of applying ARG inference to real data is 
understudied. In particular, there is a critical lack of software 
infrastructure to support evaluation and quality control of 
inferred ARGs.

@@

\section{Supplementary Material}
\subsection{Simulation of truth dataset (Fig. \ref{fig:Figure_2})}
\label{subsec:sweep_simulation}
Ancestral
histories of 300 samples were simulated with \texttt{SweepGenicSelection}
function in \texttt{msprime (version 1.3.3)}. A combination of models was used:
in the recent past, a selective sweep was simulated with a beneficial allele
situated in the middle of a 5 Mb sequence. The frequency of the allele in the
population was set at 0.0001 at the beginning of the sweep. The allele fixed in
the population at a frequency of 0.9999. The strength of selection was set
using the selection coefficient, \textit{s = 0.25}. A time increment,
\textit{dt = 1e-6} was used to step through the sweep. Mutations were added to
the ARG at a rate of 1e-8 per base pair per generation. A recombination rate of 1e-8 per base pair per generation. For simulating history before occurrence of the
sweep, a standard coalescent model (Hudson's algorithm \citep{Hudson1983}) was
used until coalescence was achieved.

@@

\clearpage
\section{Supplementary Figures}
\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{figures/SuppFig1.png}
    \caption{\textbf{\texttt{tsbrowse} applied to inference methods}
    A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate,
ARG-Needle, Relate and SINGER inferences of the truth dataset
simulated under a selective sweep model (in Figure 2 of the main text). For
SINGER, an examplar ARG from the set of output posterior ARG samples is shown.
The X coordinate represents genomic position, each horizontal segment on the
plot shows the genomic coordinates that the edge spans, and Y coordinate shows
time of either the parent or child node in the edge.}
    \label{fig:Supplementary_Figure_1}
\end{figure}

>>>>

\author[1,$\ast$]{Savita Karthikeyan \ORCID{0000-0002-4798-5746}}
\author[1]{Ben Jeffery \ORCID{/0000-0002-1982-6801}}
\author[1]{Duncan Mbuli-Robertson \ORCID{0000-0002-1660-2415}}
\author[1]{Jerome Kelleher \ORCID{0000-0002-7894-5253}}

@@

\section{Introduction}
Ancestral recombination graphs (ARGs) describe how a set of sample
sequences relate to each other at each position along the genome in a
recombining
species, and are currently the subject of intense research
interest~\citep{brandt2024promise,lewanski2024era,nielsen2024inference,
    wong2024general}. ARGs are a fundamental object in population genetics,
and although they have been of theoretical interest for
decades~\citep{Hudson1983,Griffiths1997} it is only with recent breakthroughs
in inference methods~\citep{rasmussen2014genome,
    speidel2019method,kelleher2019inferring,wohns2022unified,zhang2023biobank,
    gunnarsson2024scalable,deng2024robust} that practical application has
become possible. Varied applications have been recently proposed,
such as inferring selection~\citep{stern2019approximate,hejase2022deep} and
the spatial location of genetic ancestors~\citep{osmond2024estimating,deraje2024inferring,grundler2025geographic},
more powerful approaches to quantifying genetic
relatedness~\citep{fan2022genealogical,zhang2023biobank,gunnarsson2024scalable,lehmann2025on}
and other methodological improvements for genome wide association
studies~\citep{nowbandegani2023extremely,link2023tree},
and the development of machine learning methods using inferred
ARGs as input~\citep{hejase2022deep,pearson2023local,korfmann2024simultaneous,
whitehouse2024tree}.
While these developements are exciting, the performance 
of these new methods depends critically on the accuracy of the inferred ARGs.
Although studies benchmarking the various inference methods on 
simulated data have emerged 
\citep{brandt2022evaluation,deng2024robust,peng2024evaluating},
the practicalities of applying ARG inference to real data are 
understudied. In particular, there is a critical lack of software 
infrastructure to support evaluation and quality control of 
inferred ARGs.

@@

\section{Supplementary Methods}
\subsection{Simulation of truth dataset (Fig. \ref{fig:Figure_2})}
\label{subsec:sweep_simulation}
Ancestral
histories of 300 samples were simulated with \texttt{SweepGenicSelection}
function in \texttt{msprime (version 1.3.3)}. A combination of models was used:
in the recent past, a selective sweep was simulated with a beneficial allele
situated in the middle of a 5 Mb sequence. The frequency of the allele in the
population was set at 0.0001 at the beginning of the sweep. The allele fixed in
the population at a frequency of 0.9999. The strength of selection was set
using the selection coefficient, \textit{s = 0.25}. A time increment,
\textit{dt = 1e-6} was used to step through the sweep. Mutations were added to
the ARG at a rate of 1e-8 per base pair per generation. A recombination rate of 1e-8 per base pair per generation. For simulating history before occurrence of the
sweep, a standard coalescent model (Hudson's algorithm \citep{Hudson1983}) was
used until coalescence was achieved.

@@

\clearpage
\section{Supplementary Figures}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/SuppFig1.png}
    \caption{\textbf{\texttt{tsbrowse} applied to inference methods}
    A screenshot of \texttt{tsbrowse}'s Edges view for tsinfer+tsdate,
ARG-Needle, Relate and SINGER inferences of the truth dataset
simulated under a selective sweep model (in Figure 2 of the main text). For
SINGER, an examplar ARG from the set of output posterior ARG samples is shown.
The X coordinate represents genomic position, each horizontal segment on the
plot shows the genomic coordinates that the edge spans, and Y coordinate shows
time of either the parent or child node in the edge.}
    \label{fig:Supplementary_Figure_1}
\end{figure}
